---
title: "Analysis of COVID-Era Apple Mobility Data"
name: Kia Vaalavirta
date: 10/8/2022
output: html_document
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE)
```

## Setup


```{r set-parameters}
# assignment of the path/name of the full dataset
global_file_to_process <- "data/raw_data/applemobilitytrends-2022-04-12.csv"

# assignment of a single state name to a variable
state_to_analyze <- "New York"

# sanitize state names with spaces and replace them with underscores
state_no_spaces <- gsub(state_to_analyze,
                        pattern = " ",
                        replacement = "_")

core_name <- tools::file_path_sans_ext(basename(global_file_to_process))
```


```{r load-packages-functions}
# load the packages needed
library("dplyr")
library("readr")
library("ggplot2")
library("tidyr")
library("lubridate")
library("devtools")

# load the functions needed
source("code/functions/subset_to_a_state.R")
source("code/functions/cities_counties_by_state.R")
source("code/functions/convert_wide_to_long.R")
```

## Project Description

This rmarkdown file will use `ggplot2` [@ggplot2] and `dplyr` [@dplyr] packages to analyze the Apple mobility data from the time before the COVID-19 pandemic through shutdowns and continuing changes over the year.

The goal is to create a subset to a specific state, tally up its number of cities and counties with mobility data and make several plots to visualize those data subsets.

The state that this document will analyze is `r state_to_analyze`.

## Subsetting Data 

```{r subset-state} 
# subset state data
state_subset <- subset_to_a_state(
  input_file_name = global_file_to_process,
  state_name = state_to_analyze)
```

```{r make-subset-long}
# convert the subsetted data from wide to long
long_subset <- convert_wide_to_long(
  file_name_in = paste0("output/subsetted_states_wide/",
                        core_name,
                        "_",
                        state_no_spaces,
                        ".csv"))
```

```{r tallied-subset}
# tally up the numbers of cities and counties in the state that has mobility
# data available
tallied_subset <- cities_counties_by_state(
  file_name_in = paste0("output/subsetted_states_wide/",
                        core_name,
                        "_",
                        state_no_spaces,
                        ".csv"))
```

## Figures

```{r plot-tallied}
# create a grouped barplot of the tallied data to include in the knitted
# document, and save that figure to the output/figures directory as a png
tallied_plot <- ggplot(data = tallied_subset,
                       aes(x = geo_type,
                           y = n,
                           fill = transportation_type)) +
  geom_col(position = position_dodge()) +
  labs(title = paste("Number of cities and/or counties in",
                     state_to_analyze,
                    "with mobility data"),
       x = "Type of area",
       y = "Number of areas")

ggsave(plot = tallied_plot,
       filename = paste0("output/figures/",
                         core_name,
                         "_",
                         state_no_spaces,
                         "_tallied_plot.png"))

tallied_plot
```


```{r plot-time-series}
# create a line plot of relative driving mobility across the state with date
# on the x axis and mean relative mobility on the y axis
timeseries_plot <- long_subset %>%
  filter(transportation_type == "driving") %>%
  group_by(date) %>%
  summarize(mean_mobility = mean(rel_mobility)) %>%
  ggplot(aes(x = lubridate::ymd(date),
             y = mean_mobility)) +
  geom_line() +
  labs(title = paste("Statewide mean relative mobility driving levels in",
                     state_to_analyze,
                     "during COVID"),
       x = "Date",
       y = "Mean relative mobility")

ggsave(plot = timeseries_plot,
       filename = paste0("output/figures/",
                         core_name,
                         "_",
                         state_no_spaces,
                         "_timeseries_plot.png"))
timeseries_plot
```

```{r state-stacked}
stacked_state_plot <- ggplot(data = long_subset, aes(geo_type)) +
  geom_bar(aes(fill = transportation_type), width = 0.5) +
  labs(title = paste("The distribution of transportation types in",
                     state_to_analyze),
       x = "Type of area",
       y = "Type of transportation",
       fill = "Transportation type")


ggsave(plot = stacked_state_plot,
       filename = paste0("output/figures/",
                         core_name,
                         "_",
                         state_no_spaces,
                         "_state_stacked.png"))

stacked_state_plot
```

## Session Info

```{r session-info}
devtools::session_info()
```

## Sources Cited
